Assessing the Population-Level Conservation Effects of Marine Protected Areas

### 2021-03-08 documents/ovando-et-al-pop-effects-of-mpas.Rmd
small typo editing for final draft

<!-- What empirical evidence is there for these population-level effects of MPAs?  -->
This largely theoretical literature is generally based on modeling of closed populations with some fraction protected inside of MPAs. In contrast to this population paradigm used in MPA simulations, MPAs are often evaluated empirically at local scales using spatial response ratios, commonly measured as the ratio of biomass densities (weight of organisms per unit area) of species inside relative to selected control sites outside of MPAs [@lester2009;@halpern2003a;@edgar2014;@caselle2015]. These studies found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Several studies have also documented empirical evidence for the existence of adult or larval fish spillover affecting fish abundance [@halpern2009;@kay2012; @mcclanahan2000; @russ1996]. Where response ratios are available before and after MPA implementation, spatial before-after-control-impact (BACI) style studies have shown similarly clear and positive results [@thiault2019a]. These studies demonstrate the ability of MPAs to create differences between local fished and unfished areas,

What we do not have is clear evidence for the population-level effects of MPAs. Spatial "inside-vs-outside" studies rely on assuming that selected control sites serve as a measure of what would have happened in the absence of MPAs. Habitat characteristics are often used to justify the selection of particular fished sites as as counterfactuals (controls) in response ratios [@ferraro2018]. However, beyond habitat differences, the very spillover effects MPAs are often hoped to produce can negate the ability of spatial response ratio or BACI designs to accurately estimate the effects of MPAs, since these methods require that the control sites are conditionally unaffected by the treatment [@ferraro2018;@moffitt2013]. Spillover of adults or larvae from MPAs to control sites can mask conservation benefits, while displacement of fishing effort from MPAs to control sites (which is rarely addressed directly, @ferraro2018) can lead to overestimates of conservation gains caused by MPAs measured by spatial response ratios. Control sites sufficiently far from MPAs to negate both spillover of fish or larvae and concentration of the fishing fleet could be selected, but finding suitably distant sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. Since variations of spatial response ratios and BACI studies have been a primary source of evidence for the conservation effects of MPAs, this means that our empirical understanding of the population-level impacts of MPAs is surprisingly limited.

@@

In this paper, we present a paired theoretical and empirical assessment demonstrating the challenges in assessing the population-level impacts of MPAs. In 2003, a network of MPAs was established in the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) covering approximately 20% of the Islands' waters (an area of over 800 KM^2^). The network has been used as a model case study in protected area design around the world [@botsford2014]. We use data from the first 14 years of protection in a difference-in-difference (DiD) model [@angrist2009] to assess the population-level effect of a large  MPA network on a wide array of fin-fish species. Rather than relying on spatial controls, we use groups of species targeted and not-targeted by fishing pressure as our treatment and control groups. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and to demonstrate the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  Table.\@ref(tab:sim-table), and the equations of the simulation model can be viewed in the Appendix S1 Section 2. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `r n_distinct(outcomes$scientific_name)` species.

Many papers have presented simulation analyses of MPA outcomes [@fulton2015]. Our model incorporates core ecological and economic drivers of MPA performance assessed by these individual papers into a cohesive model, similar in spirit to @krueck2017. The simulation model consists of 50 patches with wrapped edges. For each simulation we first randomly pull a species and its associated life history traits from the FishLife [@thorson2017c] package. We pair these data with randomly selected values governing the characteristics of the simulation (Table.\@ref(tab:sim-table)). Key choices available to the model include parameters governing fishing pressure and MPA design. For a given simulation, the model randomly selects a fleet model and fishing effort allocation strategy. The fleet model can either be constant-catch (the fleet exerts as much effort as needed to maintain a fixed amount of catch), constant-effort (the fleet maintains a constant amount of effort over time), or open-access (fishing effort of the fleet expands and contracts in response to available profits). The total fishing effort exerted by the fleet can then be distributed in space uniformly, in proportion to spatial catch per unit effort, or in proportion to spatial profit per unit effort.

The simulation then begins to apply the fleet model to the population, and in a randomly selected year implements an MPA network. The model samples a percentage of the population's range to place in MPAs, and randomly assigns patches to MPAs either across a uniform system or preferentially on higher-quality habitat. The model then randomly selects whether fishing effort that used to operate inside the MPAs is redistributed to the areas outside the MPAs, or leaves the fishery entirely. We then continue the simulations with the MPAs in place. Each simulation is paired with a simulation identical in every way except that MPAs are not implemented (i.e., a simulated control). We then calculate the effect of the MPAs on the population as the difference in biomass densities in the simulation with MPAs relative to biomass densities in the simulation without MPAs.

These simulation results provide a library of plausible MPA effects for a range of biological and economic assumptions. One set of simulations was specifically designed to reflect the dynamics of the subset of species available in the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) monitoring data from the Channel Islands used in this study. For this set we only include species of the same genus as those targeted by fishing in the PISCO data, restrict fishing pressure such that the simulated populations are moderately to lightly exploited (since the PISCO data used here excludes deeper-water species such as boccacio, *Sebastes paucispinis*, which were overexploited at the inception of the Channel Island MPAs, and threatened invertebrates such as red abalone *Haliotis rufescens*), and cap the MPA size at 20% of the population's range [@rassweiler2012]. For each of these Channel Island style simulations, we calculated 1) the 'true' population-wide difference in biomass between the simulations with and without the MPAs and 2) the response ratio of biomass densities inside and outside the simulated MPAs. We then calculated the response ratios observed in the PISCO survey data from the Channel Islands, and matched these empirical results with simulations that produced similar response ratios after the same number of years of MPA protection. Because each simulation included measures of both response ratios and population-level effects, this process provides a library of simulations (and their associated attributes) that could have produced the types of empirical response ratios measured in the Channel Islands.

@@

Building off of @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group (Fig.\@ref(fig:raw-density)). Targeted  fin-fish species in the Channel Islands available to this study include California sheephead (*Semicossyphus pulcher*), and copper (*Sebastes caurinus*) and blue (*Sebastes mystinus*) rockfish. Non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*). Our regression estimates any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation. To account for the fact that sampling locations are not uniformly distributed across the islands, we weight the samples in our regression in proportion to the total area inside and outside of MPAs in the Channel Islands.

@@

where $d_i$ is the biomass density at observation *i*. To account for the fact that MPA effects will evolve over time, we estimate a vector of MPA effects in three-year blocks for the all years after the MPAs were implemented in 2003. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water temperature and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island $\beta_r$ with variance $\sigma_r^2$.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment ($MPA=1$) on the treated targeted group ($T=1$) (Table.\@ref(tab:coef-table)). We use a Bayesian hierarchical generalized linear model since it allows us to interpret our estimated effects probabilistically. Being a Bayesian regression, our DiD analysis produces posterior probability distributions (the probability distribution of our coefficients conditional on the data, priors, and model assumptions) of our coefficients, from which we can construct Bayesian credible intervals [@gelman2014a].

@@

Our targeted vs. non-targeted DiD regression provides an alternative approach to spatial controls for estimating population-level MPA effects. Over the first three years of implementation (2003-2006), the effects of the MPAs are unclear, with support for a small negative effect to a substantially positive effect, with much higher probability of a small positive effect (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`). Over the next six years the model estimates greater probabilities of an increasingly positive MPA effect, peaking in 2009-2011 with a median estimate of MPA effect of a `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)` increase in mean total biomass density of targeted species (90% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`) (Fig.\@ref(fig:mpa-effect)). These empirical estimates are in line with the outcomes that our simulation model suggests are plausible. However, in the subsequent years the trend reverses itself, and in 2015-2017 we once again see no clear effect of the MPAs (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`.) (Fig.\@ref(fig:mpa-effect)).

@@

```{r ci-map, include = TRUE,fig.width=width_1col, fig.asp=1.5, fig.cap="Map of study region; the Northern Channel Islands, California, USA. Shading shows binned number of PISCO sampling events across the time period of our study.", include=TRUE}

@@

```{r raw-density, fig.width = width_1col, fig.asp=asp_1col,fig.cap = "Centered and scaled trends in biomass densities of targeted (solid darker lines) and non-targeted (dashed lighter lines) fin-fish included in our study of the Channel Islands MPA network. Panel (a) shows mean trends across all sites, with shaded areas shwoing 95% confidence intervals. Bottom two panels show same trends but only including sites (b) outside and (c) inside MPAs. Black vertical dotted line shows MPA implementation in 2003."}

@@

```{r rr-plot, include = TRUE, fig.width=width_1col, fig.asp=asp_1col,,fig.cap="Simulated population effects (panel (a) in grey) that could have produced observed response ratios (panel (b) in black) in the Channel Islands. Panel (a) histograms of 90% posterior probability distributions of response ratios for targeted species (x-axis) over time (y-axis). Panel (b) Histograms of simulated population-level (pop.) effect on biomass densities matched to empirical response ratios in Panel (a). For response ratios, a percent difference value of zero indicates that biomass densities of targeted species are identical inside and outside MPAs, a percent difference value of one that biomass densities of targeted species are 100% greater inside MPAs relative to outside. For MPA population effect, a percent difference value of zero indicates that biomass densities are identical in the with- and without- MPA scenarios. A percent difference value of one indicates that biomass densities are 100% greater in the scenario with MPAs than the scenario without MPAs."}

>>>>

<!-- What empirical evidence is there for these population-level effects of MPAs?  -->
This largely theoretical literature is generally based on modeling of closed populations with some fraction protected inside of MPAs. In contrast to this population paradigm used in MPA simulations, MPAs are often evaluated empirically at local scales using spatial response ratios, commonly measured as the ratio of biomass densities (weight of organisms per unit area) of species inside relative to selected control sites outside of MPAs [@lester2009;@halpern2003a;@edgar2014;@caselle2015]. These studies found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Several studies have also documented empirical evidence for the existence of adult or larval fish spillover affecting fish abundance [@halpern2009;@kay2012; @mcclanahan2000; @russ1996]. Where response ratios are available before and after MPA implementation, spatial before-after-control-impact (BACI) style studies have shown similarly clear and positive results [@thiault2019a]. These studies demonstrate the ability of MPAs to create differences between local fished and unfished areas.

What we do not have is clear evidence for the population-level effects of MPAs. Spatial "inside-vs-outside" studies rely on an assumption that selected control sites serve as a measure of what would have happened in the absence of MPAs. Habitat characteristics are often used to justify the selection of particular fished sites as as counterfactuals (controls) in response ratios [@ferraro2018]. However, beyond habitat differences, the very spillover effects MPAs are often hoped to produce can negate the ability of spatial response ratio or BACI designs to accurately estimate the effects of MPAs, since these methods require that the control sites are conditionally unaffected by the treatment [@ferraro2018;@moffitt2013]. Spillover of adults or larvae from MPAs to control sites can mask conservation benefits, while displacement of fishing effort from MPAs to control sites (which is rarely addressed directly, @ferraro2018) can lead to overestimates of conservation gains caused by MPAs measured by spatial response ratios. Control sites sufficiently far from MPAs to negate both spillover of fish or larvae and concentration of the fishing fleet could be selected, but finding suitably distant sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. Since variations of spatial response ratios and BACI studies have been a primary source of evidence for the conservation effects of MPAs, this means that our empirical understanding of the population-level impacts of MPAs is surprisingly limited.

@@

In this paper, we present a paired theoretical and empirical assessment demonstrating the challenges in assessing the population-level impacts of MPAs. In 2003, a network of MPAs was established in the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) covering approximately 20% of the Islands' waters (an area of over 800 km^2^). The network has been used as a model case study in protected area design around the world [@botsford2014]. We use data from the first 14 years of protection in a difference-in-difference (DiD) model [@angrist2009] to assess the population-level effect of a large  MPA network on a wide array of fin-fish species. Rather than relying on spatial controls, we use groups of species targeted and not-targeted by fishing pressure as our treatment and control groups. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and to demonstrate the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  Table \@ref(tab:sim-table), and the equations of the simulation model can be viewed in the Appendix S1 Section 2. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `r n_distinct(outcomes$scientific_name)` species.

Many papers have presented simulation analyses of MPA outcomes [@fulton2015]. Our model incorporates core ecological and economic drivers of MPA performance assessed by these individual papers into a cohesive model, similar in spirit to @krueck2017. The simulation model consists of 50 patches with wrapped edges. For each simulation we first randomly pull a species and its associated life history traits from the FishLife [@thorson2017c] package. We pair these data with randomly selected values governing the characteristics of the simulation (Table \@ref(tab:sim-table)). Key choices available to the model include parameters governing fishing pressure and MPA design. For a given simulation, the model randomly selects a fleet model and fishing effort allocation strategy. The fleet model can either be constant-catch (the fleet exerts as much effort as needed to maintain a fixed amount of catch), constant-effort (the fleet maintains a constant amount of effort over time), or open-access (fishing effort of the fleet expands and contracts in response to available profits). The total fishing effort exerted by the fleet can then be distributed in space uniformly, in proportion to spatial catch per unit effort, or in proportion to spatial profit per unit effort.

The simulation then begins to apply the fleet model to the population, and in a randomly selected year implements an MPA network. The model samples a percentage of the population's range to place in MPAs, and randomly assigns patches to MPAs either across a uniform system or preferentially on higher-quality habitat. The model then randomly selects whether fishing effort that used to operate inside the MPAs is redistributed to the areas outside the MPAs, or leaves the fishery entirely. We then continue the simulations with the MPAs in place. Each simulation is paired with a simulation identical in every way except that MPAs are not implemented (i.e., a simulated control). Using these paired simulations, we calculate the effect of the MPAs on the population as the difference in biomass densities in the simulation with MPAs relative to biomass densities in the simulation without MPAs.

These simulation results provide a library of plausible MPA effects for a range of biological and economic assumptions. One set of simulations was specifically designed to reflect the dynamics of the subset of species available in the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) monitoring data from the Channel Islands used in this study. For this set we only include species of the same genus as those targeted by fishing in the PISCO data, restrict fishing pressure such that the simulated populations are moderately to lightly exploited (since the PISCO data used here excludes deeper-water species such as boccacio, *Sebastes paucispinis*, which were overexploited at the inception of the Channel Islands MPAs, and threatened invertebrates such as red abalone *Haliotis rufescens*), and cap the MPA size at 20% of the population's range [@rassweiler2012]. For each of these Channel Islands-style simulations, we calculated 1) the 'true' population-wide difference in biomass between the simulations with and without the MPAs and 2) the response ratio of biomass densities inside and outside the simulated MPAs. We then calculated the response ratios observed in the PISCO survey data from the Channel Islands, and matched these empirical results with simulations that produced similar response ratios after the same number of years of MPA protection. Because each simulation included measures of both response ratios and population-level effects, this process provides a library of simulations (and their associated attributes) that could have produced the types of empirical response ratios measured in the Channel Islands.

@@

Building off of @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group (Fig.\@ref(fig:raw-density)). Targeted  fin-fish species in the Channel Islands available to this study include California sheephead (*Semicossyphus pulcher*), and copper and blue rockfish (*Sebastes caurinus* and *Sebastes mystinus*). Non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*). Our regression estimates any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation. To account for the fact that sampling locations are not uniformly distributed across the islands, we weight the samples in our regression in proportion to the total area inside and outside of MPAs in the Channel Islands.

@@

where $d_i$ is the biomass density at observation *i*. To account for the fact that MPA effects will evolve over time, we estimate a vector of MPA effects in three-year blocks for the all years after the MPAs were implemented in 2003. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water temperature and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island $\beta_r$ with variance $\sigma_r^2$.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment ($MPA=1$) on the treated targeted group ($T=1$) (Table\@ref(tab:coef-table)). We use a Bayesian hierarchical generalized linear model since it allows us to interpret our estimated effects probabilistically. Being a Bayesian regression, our DiD analysis produces posterior probability distributions (the probability distribution of our coefficients conditional on the data, priors, and model assumptions) of our coefficients, from which we can construct Bayesian credible intervals [@gelman2014a].

@@

Our targeted vs. non-targeted DiD regression provides an alternative approach to spatial controls for estimating population-level MPA effects. Over the first three years of implementation (2003-2006), the effects of the MPAs are unclear, with support for a small negative effect to a substantially positive effect, with much higher probability of a small positive effect (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`). Over the next six years the model estimates greater probabilities of an increasingly positive MPA effect, peaking in 2009-2011 with a median estimate of MPA effect of a `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)` increase in mean total biomass density of targeted species (90% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`) (Fig.\@ref(fig:mpa-effect)). These empirical estimates are in line with the outcomes that our simulation model suggests are plausible. However, in the subsequent years the trend reverses, and in 2015-2017 we once again see no clear effect of the MPAs (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`.) (Fig.\@ref(fig:mpa-effect)).

@@

```{r ci-map, include = TRUE,fig.width=width_1col, fig.asp=1.5, fig.cap="Map of the study region in the Northern Channel Islands, California, USA. Shading shows binned number of PISCO sampling events across the time period of our study.", include=TRUE}

@@

```{r raw-density, fig.width = width_1col, fig.asp=asp_1col,fig.cap = "Centered and scaled trends in biomass densities of targeted (solid darker lines) and non-targeted (dashed lighter lines) fin-fish included in our study of the Channel Islands MPA network. Panel (a) shows mean trends across all sites, with shaded areas showing 95% confidence intervals. Bottom two panels show same trends but only including sites (b) outside and (c) inside MPAs. Black vertical dotted line shows MPA implementation in 2003."}

@@

```{r rr-plot, include = TRUE, fig.width=width_1col, fig.asp=asp_1col,,fig.cap="Simulated population effects (panel (b) in grey) that could have produced empirically-observed response ratios (panel (a) in black) in the Channel Islands. Panel (a) histograms of 90% posterior probability distributions of response ratios for targeted species (x-axis) over time (y-axis). Panel (b) Histograms of simulated population-level effects on biomass densities matched to empirical response ratios in Panel (a). For response ratios, a percent difference value of zero indicates that biomass densities of targeted species are identical inside and outside MPAs, a percent difference value of one that biomass densities of targeted species are 100% greater inside MPAs relative to outside. For MPA population effect, a percent difference value of zero indicates that biomass densities are identical in the with- and without- MPA scenarios. A percent difference value of one indicates that biomass densities are 100% greater in the scenario with MPAs than the scenario without MPAs."}
